Quantum Algorithms for Portflolio Optimisation and Option pricing

### 2024-06-04 project/paper-v2.tex
update project paper VQE

\title{Quantum Algorithms for Portflolio Optimisation in Finance}

@@

The purpose of option pricing is to estimate Eq. \ref{eq:8} at the expiration date and then, determine the \textbf{fair value}, which indicates the amount of money an 
investor should pay to enter the option contract today, given the current market conditions. 
The challenging part is to compute the expected value of the option payoff, which is usually done using Monte Carlo methods. In section \ref{sec:literature2} we will review how this problem can be solved using quantum algorithms.

\subsubsection{Variational Quantum Algorithms}
Variational Quantum Algorithms (VQAs) are hybrid quantum-classical algorithms that 
are well suited for NISQ devices and allow us to get approximate solutions to optimisation and combinatorial problems efficiently \cite{Cerezo2021}. This algorithms rely on the fact that we can, in 
principle, reformulate any optimisation problem as a QUBO problem \cite{Glover2019}.

@@

The starting point is to define a cost function
\begin{equation}
  \label{eq:9}
  C(\boldsymbol{\theta}) =\langle\psi(\boldsymbol{\theta})|H|\psi(\boldsymbol{\theta})\rangle
\end{equation}
where $H$ is the Hamiltonian of our problem, $\boldsymbol{\theta}$ is a set of trainable parameters, and a trial state $|\psi(\boldsymbol{\theta})\rangle = U(\boldsymbol{\theta}) |0\rangle ^{\otimes n}$ for some ansatz $U(\boldsymbol{\theta})$

\begin{figure*}[ht!]
  \begin{center} 
  \includegraphics[width=0.85\textwidth]{QAOA-Blekos.pdf}
  \caption{\label{fig:qaoa} Quantum Approximate Optimisation Algorithm (QAOA) algorithm. Adapted from \cite{Blekos2024}.}
  \end{center}
\end{figure*}

>>>>

\title{Quantum Algorithms for Portflolio Optimisation and Option pricing}

@@

The purpose of option pricing is to estimate Eq. \ref{eq:8} at the expiration date and then, determine the \textbf{fair value}, which indicates the amount of money an 
investor should pay to enter the option contract today, given the current market conditions. 
The challenging part is to compute the expected value of the option payoff, which is usually done using Monte Carlo methods. In section \ref{sec:literature2} we will discuss the quantum approach to 
this problem.

\subsection{Variational Quantum Algorithms}
Variational Quantum Algorithms (VQAs) are hybrid quantum-classical algorithms that 
are well suited for NISQ devices and allow us to get approximate solutions to optimisation and combinatorial problems efficiently \cite{Cerezo2021}. This algorithms rely on the fact that we can, in 
principle, reformulate any optimisation problem as a QUBO problem \cite{Glover2019}.

@@

The starting point is to define a cost function
\begin{equation}
  \label{eq:9}
  C(\boldsymbol{\theta}) =\langle\psi(\boldsymbol{\theta})|H|\psi(\boldsymbol{\theta})\rangle
\end{equation}
where $H$ is the Hamiltonian of our problem, that can be written as the weighted sum of Pauli operators \cite{Tilly2022}, such
\begin{figure*}[ht!]
  \begin{center} 
  \includegraphics[width=0.85\textwidth]{QAOA-Blekos.pdf}
  \caption{\label{fig:qaoa} Quantum Approximate Optimisation Algorithm (QAOA). Adapted from \cite{Blekos2024}.}
  \end{center}
\end{figure*}
that: 
\begin{equation}
  \label{eq:10}
  \mathcal{H} = \sum_{i} c_i P_i
\end{equation}
where $P_i \in \{I, X, Y, Z\}^{\otimes N}$ are called Pauli strings and $N$ is the number of qubits, and $c_i$ are the coefficients of the Pauli strings.  Additionally, $\boldsymbol{\theta}$ is a set of trainable parameters for some trial state $|\psi(\boldsymbol{\theta})\rangle = U(\boldsymbol{\theta}) |0\rangle ^{\otimes n}$. So, Eq. \ref{eq:9} becomes:
\begin{equation}
  \label{eq:11}
  C(\boldsymbol{\theta}) =\sum_i c_i \langle \boldsymbol{0}|U(\boldsymbol{\theta})^\dagger P_i U(\boldsymbol{\theta})|\boldsymbol{0}\rangle
\end{equation}
for some ansatz $U(\boldsymbol{\theta})$, that can be written as
\begin{equation}
  \label{eq:12}
  U(\boldsymbol{\theta}) = \prod_{k=1}^{p} e^{-i\theta_k \mathcal{H}_k}
\end{equation}
where $\mathcal{H}_k$ are the terms of the Hamiltonian $\mathcal{H}$  and $p$ determines the precision of the approximation \cite{Cerezo2021}.

Therefore, the VQE consists of classical algorithm that uses a quantum algorithm as a subroutine. We first initialise the state 
$|\psi(\boldsymbol{\theta})\rangle$ with some initial parameters $\boldsymbol{\theta}$, then we measure Eq. \ref{eq:11} and use a classical optimisation algorithm to update 
the parameters $\boldsymbol{\theta}$ until we reach a minimum value of $C(\boldsymbol{\theta})$.

The variational principle ensures that at some point the VQE algorithm will converge to:
\begin{equation}
  \label{eq:13}
  E_0 \leq \frac{\langle \psi(\boldsymbol{\theta})|\mathcal{H}|\psi(\boldsymbol{\theta})\rangle}{\langle \psi(\boldsymbol{\theta})|\psi(\boldsymbol{\theta})\rangle}
\end{equation}
where $E_0$ is the ground state energy of the Hamiltonian $\mathcal{H}$ and $|\psi(\boldsymbol{\theta})\rangle$ is some trial state.

In a similar fashion, the Quantum Approximate Optimisation Algorithm (QAOA) is another VQA that was first 
conceived as
